# Comparison of two PCA packages

-   Datasets: `decathlon2`

-   Algorithms:

    -   PCA

## prcomp vs princomp

<http://www.sthda.com/english/articles/31-principal-component-methods-in-r-practical-guide/118-principal-component-analysis-in-r-prcomp-vs-princomp/>

## General methods for principal component analysis

There are two general methods to perform PCA in R :

-   Spectral decomposition which examines the covariances / correlations between variables
-   Singular value decomposition which examines the covariances / correlations between individuals

The function princomp() uses the spectral decomposition approach. The functions prcomp() and PCA()[FactoMineR] use the singular value decomposition (SVD).

### prcomp() and princomp() functions

The simplified format of these 2 functions are :

    prcomp(x, scale = FALSE)
    princomp(x, cor = FALSE, scores = TRUE)

1.  Arguments for prcomp():\
    `x`: a numeric matrix or data frame\
    `scale`: a logical value indicating whether the variables should be scaled to have unit variance before the analysis takes place

2.  Arguments for princomp():\
    `x`: a numeric matrix or data frame `cor`: a logical value. If TRUE, the data will be centered and scaled before the analysis `scores`: a logical value. If TRUE, the coordinates on each principal component are calculated

### Loading factoextra

```{r eval=FALSE}
# install.packages("factoextra")
```

```{r}
library(factoextra)
```

### Loading the decathlon dataset

We'll use the data sets `decathlon2` [in factoextra], which has been already described at: PCA - Data format.

Briefly, it contains:

-   Active individuals (rows 1 to 23) and active variables (columns 1 to 10), which are used to perform the principal component analysis
-   Supplementary individuals (rows 24 to 27) and supplementary variables (columns 11 to 13), which coordinates will be predicted using the PCA information and parameters obtained with active individuals/variables.

```{r}
library("factoextra")
data(decathlon2)
decathlon2.active <- decathlon2[1:23, 1:10]
head(decathlon2.active[, 1:6])
```

```{r}
decathlon2.supplementary <- decathlon2[24:27, 1:10]
head(decathlon2.supplementary[, 1:6])
```

## Compute PCA in R using prcomp()

In this section we'll provide an easy-to-use R code to compute and visualize PCA in R using the prcomp() function and the factoextra package.

\`. Load factoextra for visualization

```{r}
library(factoextra)
```

2.  compute PCA

```{r}
# compute PCA
res.pca <- prcomp(decathlon2.active, scale = TRUE)
```

3.  Visualize eigenvalues (scree plot). Show the percentage of variances explained by each principal component.

```{r}
# Visualize eigenvalues (scree plot).
fviz_eig(res.pca, addlabels = TRUE, ylim = c(0, 50))
```

> From the plot above, we might want to stop at the fifth principal component. 87% of the information (variances) contained in the data are retained by the first five principal components.

## Plots: quality and contribution

4.  Graph of individuals. Individuals with a similar profile are grouped together.

```{r}
# Graph of individuals.
fviz_pca_ind(res.pca,
             col.ind = "cos2", # Color by the quality of representation
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE     # Avoid text overlapping
             )
```

5.  Graph of variables. Positive correlated variables point to the same side of the plot. Negative correlated variables point to opposite sides of the graph.

```{r}
# Graph of variables.
fviz_pca_var(res.pca,
             col.var = "contrib", # Color by contributions to the PC
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE     # Avoid text overlapping
             )
```

6.  Biplot of individuals and variables

```{r}
# Biplot of individuals and variables
fviz_pca_biplot(res.pca, repel = TRUE,
                col.var = "#2E9FDF", # Variables color
                col.ind = "#696969"  # Individuals color
                )
```

## Access to the PCA results

```{r}
library(factoextra)
# Eigenvalues
eig.val <- get_eigenvalue(res.pca)
eig.val
  
# Results for Variables
res.var <- get_pca_var(res.pca)
res.var$coord          # Coordinates
res.var$contrib        # Contributions to the PCs
res.var$cos2           # Quality of representation 
# Results for individuals
res.ind <- get_pca_ind(res.pca)
res.ind$coord          # Coordinates
res.ind$contrib        # Contributions to the PCs
res.ind$cos2           # Quality of representation 
```


## Theory behind PCA results

### PCA results for variables

Here we'll show how to calculate the PCA results for variables: coordinates, cos2 and contributions:

`var.coord` = loadings \* the component standard deviations `var.cos2` = var.coord\^2 `var.contrib`. The contribution of a variable to a given principal component is (in percentage) : (var.cos2 \* 100) / (total cos2 of the component)

```{r}
# Helper function 
#::::::::::::::::::::::::::::::::::::::::
var_coord_func <- function(loadings, comp.sdev){
  loadings*comp.sdev
}
```

```{r}
# Compute Coordinates
#::::::::::::::::::::::::::::::::::::::::
loadings <- res.pca$rotation
sdev <- res.pca$sdev
var.coord <- t(apply(loadings, 1, var_coord_func, sdev)) 
head(var.coord[, 1:4])
```

```{r}
# Compute Cos2
#::::::::::::::::::::::::::::::::::::::::
var.cos2 <- var.coord^2
head(var.cos2[, 1:4])
```

```{r}
# Compute contributions
#::::::::::::::::::::::::::::::::::::::::
comp.cos2 <- apply(var.cos2, 2, sum)
contrib <- function(var.cos2, comp.cos2){var.cos2*100/comp.cos2}
var.contrib <- t(apply(var.cos2,1, contrib, comp.cos2))
head(var.contrib[, 1:4])
```

### PCA results for individuals

-   `ind.coord` = res.pca\$x

-   Cos2 of individuals. Two steps:

    -   Calculate the square distance between each individual and the PCA center of gravity: d2 = [(var1_ind_i - mean_var1)/sd_var1]\^2 + ...+ [(var10_ind_i - mean_var10)/sd_var10]\^2 + ...+..
    -   Calculate the cos2 as ind.coord\^2/d2

-   Contributions of individuals to the principal components: 100 \* (1 / number_of_individuals)\*(ind.coord\^2 / comp_sdev\^2). Note that the sum of all the contributions per column is 100

```{r}
# Coordinates of individuals
#::::::::::::::::::::::::::::::::::
ind.coord <- res.pca$x
head(ind.coord[, 1:4])
```

```{r}
# Cos2 of individuals
#:::::::::::::::::::::::::::::::::
# 1. square of the distance between an individual and the
# PCA center of gravity
center <- res.pca$center
scale<- res.pca$scale

getdistance <- function(ind_row, center, scale){
  return(sum(((ind_row-center)/scale)^2))
}

d2 <- apply(decathlon2.active,1, getdistance, center, scale)
# 2. Compute the cos2. The sum of each row is 1
cos2 <- function(ind.coord, d2){return(ind.coord^2/d2)}
ind.cos2 <- apply(ind.coord, 2, cos2, d2)
head(ind.cos2[, 1:4])
```

```{r}
# Contributions of individuals
#:::::::::::::::::::::::::::::::
contrib <- function(ind.coord, comp.sdev, n.ind){
  100*(1/n.ind)*ind.coord^2/comp.sdev^2
}
ind.contrib <- t(apply(ind.coord, 1, contrib, 
                       res.pca$sdev, nrow(ind.coord)))
head(ind.contrib[, 1:4])
```
